---
title: "01 batch correction"
author: "Madeline Krieger"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    theme: "cosmo"
    code_folding: "show"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE,
                      error = FALSE,
                      fig.path = "./figures/01/",
                      fig.keep = "all",
                      dev = c("png"),
                      cache = FALSE,
                      # Do not use lazy loading which fails for large objects
                      cache.lazy = FALSE)
```


# 1.0 Load and Define

## 1.1 Load packages and presets
### 1.1.1 Outpside Packages
```{r 1.1_load-packages}
library("pacman")
pacman::p_load("rmarkdown", "tidyverse", "phyloseq", "knitr","kableExtra", "here", "plyr", "ggpubr", "microViz", "readxl", "microbiome", "pheatmap", "vegan", "DirichletMultinomial", "reshape2", "magrittr", "microshades", "pheatmap","vegan", "data.table", "Polychrome", "fantaxtic","cetcolor", "topicmodels", "ldatuning", "cowplot", "MicrobiomeStat", "randomForest", "caret", "mlbench", "MLmetrics", "mia", "here", "patchwork", "digest")

here::i_am("01b_batch_correction.Rmd")

``` 


### 1.1.2 My packages
```{r}
source("MK_Microbiome_Functions.R")
```



### 1.1.3 Colors and presets
```{r}
colors_all= c("Abscess" = "#FF495C", "Tumor"="#F0C808","Plaque"="#083D77", "AN"="#43AA8B" )
colors_study= c("AdultAbscess" = "#FF495C", "Pediatric"="#3399FF","OSCC"="#03CEA4" )


core_colors = c("#F76F8E","#083D77","#B8D4E3", "#FAAA00", "#3399FF","#FF495C","#477071", '#03CEA4',  "#5F00BA", "#BDAC9E", "white", "#FFD900")


resolution <- c("Phylum", "Class", "Order", "Family", "Genus", "Species")

set.seed(1234)
date = Sys.Date()
```


## 1.3 Load and process

### 1.3.1 Load Data from 00
```{r}
# Load the saved Rds file
loaded_objects <- readRDS("00data.rds")

# Assign back the objects to the environment, if needed
list2env(loaded_objects, .GlobalEnv)
```


### 1.3.2 Process Phyloseq
```{r}
process_phyloseq(merged_phylo_obj_fs)
```


# 2.0 Ordinations: Comparing Studies

## 2.1 PCA 
```{r PCA, fig.width=10, fig.height=15}
plots <- lapply(resolution, function(rank) {
  plot_pca(phyloseq_obj = merged_phylo_obj_f, 
           rank_transformation = rank, 
           variable = "Study", 
           colors_list = colors_study)})

combined_plot <- wrap_plots(plots, ncol = 2) & theme(legend.position = "bottom") 
combined_plot
```


## 2.2 PCoA NMDS Bray 
```{r PCoA_bray, fig.width=10, fig.height=15}
plots <- lapply(resolution, function(rank) {
  plot_PCoA(phyloseq_obj = merged_phylo_obj_f,
    rank_transformation = rank,
    trans_type = "identity",        
    dist_cal_type = "bray",   
    ord_calc_method = "NMDS",
    variable = "Study", 
    colors_list = colors_study)})

combined_plot <- wrap_plots(plots, ncol = 2) & theme(legend.position = "bottom") 
print(combined_plot)
```


## 2.3 PCoA NMDS Jaccard 
```{r PCoA_jaccard, fig.width=10, fig.height=15}
plots <- lapply(resolution, function(rank) {
  plot_PCoA(phyloseq_obj = merged_phylo_obj_f,
    rank_transformation = rank,
    trans_type = "identity",        
    dist_cal_type = "jaccard",   
    ord_calc_method = "NMDS",
        variable = "Study", 
    colors_list = colors_study)})

combined_plot <- wrap_plots(plots, ncol = 2) & theme(legend.position = "bottom") 
print(combined_plot)
```



# 3.0 MBCES Batch correction
https://bioconductor.org/packages/release/bioc/vignettes/MBECS/inst/doc/mbecs_vignette.html
```{r}
library(MBECS)
```

## 3.1 Pre-processing
### 3.1.1 Create object
```{r}
run_MBCES <- function(ps_obj, clr_methods=c("bat", "svd", "pn", "bmc"), tss_methods=c( "svd", "pn", "bmc")){
  
  # Create a unique hash for the parameters
  param_hash <- digest(list(ps_obj, clr_methods, tss_methods))
  # File path for the analysis results
  result_file <- file.path("saved_analysis_files/", paste0("mbec.obj_", param_hash, ".rds"))

  if (file.exists(result_file)) {
      message("Analysis already run. Loading results...")
      mbec.obj <- readRDS(result_file)
    } else {
      message("Running analysis...")
    
    #Create Object
    mbec.obj <- mbecProcessInput(ps_obj, 
                             required.col = c("Study", "condition"))
  
    #Apply Transformations
    mbec.obj <- mbecTransform(mbec.obj, method = "tss")
    mbec.obj <- mbecTransform(mbec.obj, method = "clr", offset = 0.0001)
    
    #CLR prelim report
    mbecReportPrelim(input.obj=mbec.obj, model.vars=c("Study", "condition"), type="clr")
 
    #TSS prelim report
    mbecReportPrelim(input.obj=mbec.obj, model.vars=c("Study", "condition"), type="tss")

    #Run CLR
    mbec.obj<- mbecRunCorrections(mbec.obj, model.vars=c("Study", "condition"), 
                           method= clr_methods,
                           type = "clr")
    
    #Run TSS
    mbec.obj<- mbecRunCorrections(mbec.obj, model.vars=c("Study", "condition"), 
                           method= tss_methods , #got an error when running "bat"
                           type = "tss")
    
    #CLR Post-report
    mbecReportPost(input.obj=mbec.obj, model.vars=c("Study", "condition"), 
               type="clr")
    
    #TSS Post-report
    mbecReportPost(input.obj=mbec.obj, model.vars=c("Study", "condition"), 
               type="tss")
    
      # Save the results
    saveRDS(mbec.obj, result_file)
    }
  
    return(mbec.obj)
}

mbec.obj<- run_MBCES(merged_phylo_obj_fs)
```

### 3.1.2 Extract data
```{r}
ps.tss.pn <- mbecGetPhyloseq(mbec.obj, type="tss", label="pn")
ps.clr.pn <- mbecGetPhyloseq(mbec.obj, type="clr", label="pn")

ps.tss.bmc <- mbecGetPhyloseq(mbec.obj, type="tss", label="bmc")
ps.clr.bmc <- mbecGetPhyloseq(mbec.obj, type="clr", label="bmc")

ps.tss.svd <- mbecGetPhyloseq(mbec.obj, type="tss", label="svd")
ps.clr.svd <- mbecGetPhyloseq(mbec.obj, type="clr", label="svd")

```

## 3.2 Compare Data: PCA

### 3.2.1 TSS PN
```{r, fig.width=10, fig.height=15}
plots <- lapply(resolution, function(rank) {
  plot_pca(phyloseq_obj = merged_phylo_obj_fs, 
           rank_transformation = rank, 
           variable = "Study", 
           colors_list = colors_study)})

combined_plot <- wrap_plots(plots, ncol = 2) & theme(legend.position = "bottom") 
combined_plot
```

### 3.2.2 TSS BMC
```{r, fig.width=10, fig.height=15}
plots <- lapply(resolution, function(rank) {
  plot_pca(phyloseq_obj = ps.tss.bmc, 
           rank_transformation = rank, 
           variable = "Study", 
           colors_list = colors_study)})

combined_plot <- wrap_plots(plots, ncol = 2) & theme(legend.position = "bottom") 
combined_plot
```

### 3.2.3 TSS SVD
```{r, fig.width=10, fig.height=15}
plots <- lapply(resolution, function(rank) {
  plot_pca(phyloseq_obj = ps.tss.svd, 
           rank_transformation = rank, 
           variable = "Study", 
           colors_list = colors_study)})

combined_plot <- wrap_plots(plots, ncol = 2) & theme(legend.position = "bottom") 
combined_plot
```


# 4.0 ConqR Batch Correction
Can't use because I don't have enough metadata elements....it wants more than one covariate 
```{r}
library(ConQuR)
library(doParallel)
```




