---
title: "02a Pediatric"
author: "Madeline Krieger"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    theme: "cosmo"
    code_folding: "show"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE,
                      error = FALSE,
                      fig.path = "./figures/02a/",
                      fig.keep = "all",
                      dev = c("png"),
                      cache = FALSE,
                      # Do not use lazy loading which fails for large objects
                      cache.lazy = FALSE)
```


# 1.0 Load and Define

## 1.1 Load packages and presets
### 1.1.1 Packages
```{r 1.1_load-packages}
library("pacman")
pacman::p_load("rmarkdown", "tidyverse", "phyloseq", "knitr","kableExtra", "here", "plyr", "ggpubr", "microViz", "readxl", "microbiome", "pheatmap", "vegan", "DirichletMultinomial", "reshape2", "magrittr", "microshades", "pheatmap","vegan", "data.table", "Polychrome", "fantaxtic","cetcolor", "topicmodels", "ldatuning", "cowplot", "MicrobiomeStat", "randomForest", "caret", "mlbench", "MLmetrics", "mia", "here", "patchwork", "digest", "ANCOMBC", "Maaslin2")

here::i_am("02a_Pediatric.Rmd")

``` 


### 1.1.2 My packages
```{r}
load("MK_Microbiome_Functions.RData")
```



### 1.1.2 Colors and presents
```{r}
colors_all= c("Abscess" = "#FF495C", "Tumor"="#F0C808","Plaque"="#083D77", "AN"="#43AA8B" )
plot_colors= c("Abscess" = "#FF495C", "Plaque"="#083D77","abscess" = "#FF495C", "plaque"="#083D77", 
               "abnormal" = "#FF495C", "control"="#083D77",  "Abnormal" = "#FF495C", "Control"="#083D77"  )


core_colors = c("#F76F8E","#083D77","#B8D4E3", "#FAAA00", "#3399FF","#FF495C","#477071", '#03CEA4',  "#5F00BA", "#BDAC9E", "white", "#FFD900")

maaslin2_colors= c("CLR_LOG" = "#2c45b5", "CLR_NONE"="#86B8FD", 
                   "CSS_LOG" = "#FF495C", "CSS_NONE"="#F2929A",
                   "TMM_LOG" = "#F4B701", "TMM_NONE"="#FEE59A",
                   "TSS_LOG" = "#8FD694", "TSS_NONE"="#D2EED4",
                   "TSS_LOGIT" = "#354F52", "TSS_AST"="#CEDDDF")

pal20 <- createPalette(22, c("#F76F8E","#03CEA4", "#083D77"), range = c(30, 80))
pal20 <- unname(as.list(pal20))

resolution <- c("Phylum", "Class", "Order", "Family", "Genus", "Species")

set.seed(1234)
date = Sys.Date()
```


## 1.2 Load data and process

### 1.2.1 Load Data from 00
```{r}
# Load the saved Rds file
loaded_objects <- readRDS("00data.rds")

# Assign back the objects to the environment, if needed
list2env(loaded_objects, .GlobalEnv)

ps_obj_ped_f <- subset_samples(merged_phylo_obj_f, Study == "Pediatric")
ps_obj_ped_fs <- subset_samples(merged_phylo_obj_fs, Study == "Pediatric")
```


### 1.2.2 Process Phyloseq
```{r}
process_phyloseq(ps_obj_ped_fs)
```


# 2.0 Ordinations

## 2.1 PCA 
```{r, fig.width=10, fig.height=15}
plots <- lapply(resolution, function(rank) {
  plot_pca(phyloseq_obj = ps_obj_ped_f, 
           rank_transformation = rank, 
           variable = "Type", 
           colors_list = colors_all)})

combined_plot <- wrap_plots(plots, ncol = 2) & theme(legend.position = "bottom") 
combined_plot
```


## 2.2 PCoA NMDS Bray 
```{r, fig.width=10, fig.height=15}
plots <- lapply(resolution, function(rank) {
  plot_PCoA(phyloseq_obj = ps_obj_ped_f,
    rank_transformation = rank,
    trans_type = "identity",        
    dist_cal_type = "bray",   
    ord_calc_method = "NMDS",
    variable = "Type", 
    colors_list = colors_all)})

combined_plot <- wrap_plots(plots, ncol = 2) & theme(legend.position = "bottom") 
print(combined_plot)
```


## 2.3 PCoA NMDS Jaccard 
```{r, fig.width=10, fig.height=15}
plots <- lapply(resolution, function(rank) {
  plot_PCoA(phyloseq_obj = ps_obj_ped_f,
    rank_transformation = rank,
    trans_type = "identity",        
    dist_cal_type = "jaccard",   
    ord_calc_method = "NMDS",
        variable = "Type", 
    colors_list = colors_all)})

combined_plot <- wrap_plots(plots, ncol = 2) & theme(legend.position = "bottom") 
print(combined_plot)
```



# 3.0 Alpha diversity
```{r}
a_my_comparisons <- list( c("Abscess", "Plaque"))
symnum.args = list(cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1), symbols = c("****", "***", "**", "*", "ns"))

p_Shannon <- plot_richness(ps_obj_ped_f, x="Type", measures="Shannon", color = "Type")+
  geom_boxplot(alpha=0.6)+ 
    theme_bw(base_size=14) +
  theme(legend.position="none",         
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12), 
        axis.title.x = element_blank(),
        axis.title.y = element_text(size=12), 
        strip.background = element_rect(fill = "white", color = "white"), 
        strip.text = element_text(size = 14, face = "bold"))  +
  stat_compare_means(method = "wilcox.test", comparisons = a_my_comparisons, label = "p.format")+
  scale_color_manual(values = colors_all) +
  ylim(0, 6.2)


p_Chao1 <- plot_richness(ps_obj_ped_f, x="Type", measures="Chao1", color = "Type")+
  geom_boxplot(alpha=0.6)+ 
    theme_bw(base_size=14) +
  theme(legend.position="none",         
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12), 
        axis.title.x = element_blank(),
        axis.title.y = element_text(size=12), 
        strip.background = element_rect(fill = "white", color = "white"), 
        strip.text = element_text(size = 14, face = "bold"))  +
  stat_compare_means(method = "wilcox.test", comparisons = a_my_comparisons, label = "p.format")+
  scale_color_manual(values = colors_all) +
  ylim(0, 470)


p_Observed <- plot_richness(ps_obj_ped_f, x="Type", measures="Observed", color = "Type")+
  geom_boxplot(alpha=0.6)+ 
    theme_bw(base_size=14) +
  theme(legend.position="none",         
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12), 
        axis.title.x = element_blank(),
        axis.title.y = element_text(size=12), 
        strip.background = element_rect(fill = "white", color = "white"), 
        strip.text = element_text(size = 14, face = "bold"))  +
  stat_compare_means(method = "wilcox.test", comparisons = a_my_comparisons, label = "p.format")+
  scale_color_manual(values = colors_all) +
  ylim(0, 470)

plot <- ggarrange(p_Shannon, p_Observed, p_Chao1, ncol=3, nrow=1, common.legend = TRUE, legend="bottom")
annotate_figure(plot, top = text_grob(paste0("Alpha Diversity"), color = "black", face = "bold", size = 14))
```

# 4.0 Taxa Plots

## 4.1 Average Bar Plots

### 4.1.1 Phylum 
```{r taxa_phylum, fig.height=4, fig.width=4}

top_df <- ps_obj_ped_fs_t_n %>%
  tax_glom(., "Phylum") %>%
  get_top_taxa(., 10, relative = TRUE, discard_other = TRUE) %>% psmelt(.)
top_df$Phylum <- gsub("p__","",as.character(top_df$Phylum))

top_df %>%
  dplyr::group_by(Type,Phylum)%>%
  dplyr::summarise(Average_Abundance = mean(Abundance))%>%
  ggbarplot(x= "Type", y="Average_Abundance",
            fill = "Phylum",
            panel.labs.font = list(size=12),
            panel.labs.background = list(color = NULL, fill = "white"),
            font.tickslab = 14,
            font.legend = c(10),
            palette = core_colors, 
            strip.position = "top",
            color = "black",
            x.text.angle = 45, 
            y.text.angle = 0, 
            font.x = 14,
            font.y = 12,
            font.title = 16,
            rotate = FALSE,
            title = "Phlyum Level", 
            xlab = "",
            ylab = "Average Relative Abundance",
            ggtheme = theme_bw()) + 
  font("title", face="bold") +
  theme(plot.title=element_text(hjust=0.5), plot.subtitle=element_text(hjust=0.5))
```

### 4.1.2 Genus 
```{r taxa_genus, fig.height=4, fig.width=3}

top_df <- ps_obj_ped_fs_t_n %>%
  tax_glom(., "Genus") %>%
  get_top_taxa(., 10, relative = TRUE, discard_other = TRUE) %>% psmelt(.)
top_df$Genus <- gsub("g__","",as.character(top_df$Genus))

top_df %>%
  dplyr::group_by(Type,Genus)%>%
  dplyr::summarise(Average_Abundance = mean(Abundance))%>%
  ggbarplot(x= "Type", y="Average_Abundance",
            fill = "Genus",
            panel.labs.font = list(size=12),
            panel.labs.background = list(color = NULL, fill = "white"),
            font.tickslab = 14,
            font.legend = c(10),
            palette = core_colors, 
            strip.position = "top",
            color = "black",
            x.text.angle = 45, 
            y.text.angle = 0, 
            font.x = 14,
            font.y = 12,
            font.title = 16,
            rotate = FALSE,
            title = "Top 10 Genera", 
            xlab = "",
            ylab = "Average Relative Abundance",
            ggtheme = theme_bw()) + 
  font("title", face="bold") +
  theme(plot.title=element_text(hjust=0.5), plot.subtitle=element_text(hjust=0.5), legend.text = element_text(face = "italic"))
```

### 4.1.3 Microshades Genus MB
```{r, fig.width=8, fig.height=4}
# Use microshades function prep_mdf to agglomerate, normalize, and melt the phyloseq object
mdf_prep <- prep_mdf(ps_obj_ped_fs)

mdf_prep <- mdf_prep %>%
  mutate(Kingdom = gsub("k__", "", Kingdom)) %>%
  mutate(Phylum = gsub("p__", "", Phylum)) %>%
  mutate(Class = gsub("c__", "", Class)) %>%
  mutate(Order = gsub("o__", "", Order)) %>%
  mutate(Family = gsub("f__", "", Family)) %>%
  mutate(Genus = gsub("g__", "", Genus)) %>%
  mutate(Species = gsub("s__", "", Species)) 

print(unique(mdf_prep$Phylum))

# Create a color object for the specified data
color_obj <- create_color_dfs(mdf_prep, 
            selected_groups = c("Proteobacteria", "Actinobacteria","Bacteroidetes", "Fusobacteria",
                                "Spirochaetes"),
    cvd = FALSE, top_orientation=TRUE)

# Extract
mdf <- color_obj$mdf
cdf <- color_obj$cdf

plot_1 <- plot_microshades(mdf, cdf)

plot_1 + scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme_bw() +
  theme(legend.key.size = unit(0.2, "cm"), text=element_text(size=10), ) +
  theme(axis.text.x = element_text(size= 6)) +
  facet_wrap(~Type, scales= "free_x") 
```



### 4.1.4 Species 
```{r species}
top_df <- ps_obj_ped_fs_t_n %>%
  tax_glom(., "Species") %>%
  get_top_taxa(., 10, relative = TRUE, discard_other = TRUE) %>% psmelt(.)
top_df$Species <- gsub("s__","",as.character(top_df$Species))

top_df %>%
  dplyr::group_by(Type,Species)%>%
  dplyr::summarise(Average_Abundance = mean(Abundance))%>%
  ggbarplot(x= "Type", y="Average_Abundance",
            fill = "Species",
            panel.labs.font = list(size=12),
            panel.labs.background = list(color = NULL, fill = "white"),
            font.tickslab = 10,
            font.legend = c(10),
            palette = pal20, 
            strip.position = "top",
            color = "black",
            x.text.angle = 90, 
            y.text.angle = 0, 
            font.x = 12,
            font.y = 12,
            font.title = 16,
            rotate = TRUE,
            title = "", 
            subtitle = "Species level",
            xlab = "Condition",
            ylab = "Average Abundance",
            ggtheme = theme_bw()) + 
  font("title", face="bold") +
  theme(plot.title=element_text(hjust=0.5), plot.subtitle=element_text(hjust=0.5))
```


## 4.2 Per taxa plots

### 4.2.1 Genus TSS Dot
```{r, fig.width=15, fig.height=50}
plot_all_taxa(ps_obj_ped_fs, "TSS", "condition", "Genus", "dot", plot_colors)
```


# 5.0 Core Microbiome
```{r}
core_ps <- ps_obj_ped_fs_t_n

# Agglomerate to the species level
core_ps_species <- tax_glom(core_ps, taxrank = "Species")
core_ps_genus <- tax_glom(core_ps, taxrank = "Genus")

# Initialize an empty dataframe to store results
results <- data.frame(prevalence = numeric(), percent_OTUs = numeric())

# Get the total number of OTUs at the species level
total_OTUs <- length(taxa_names(core_ps_species))

# Loop through prevalence values from 0.1 to 1
for (p in seq(0.1, 1, by = 0.1)) {
  core.taxa <- core_members(core_ps_species, detection = 0.00001, prevalence = p)
  num_OTUs <- length(core.taxa)
  percent_OTUs <- (num_OTUs / total_OTUs) * 100
  
  # Append the results to the dataframe
  results <- rbind(results, data.frame(prevalence = p*100, percent_OTUs = percent_OTUs))
}

# Create a line graph using ggplot2
ggplot(results, aes(x = prevalence, y = percent_OTUs)) +
  geom_line() +
  geom_point() +
  labs(title = "Percent of Total Species Detectioned",
       x = "Percent of Samples Containing Taxa",
       y = "Percent of Total Taxa") +
  theme_minimal()
```

```{r, fig.height=9, fig.width=7}

# Get the total list of species
genus_list <- taxa_names(core_ps_genus) 

# Initialize an empty list to store core genera at each prevalence level
core_genus_list <- list()

# Loop through prevalence values from 0.1 to 1
for (p in seq(0.1, 1, by = 0.05)) {
  core.taxa <- core_members(core_ps_genus, detection = 0.00001, prevalence = p)
  core_genus_list[[as.character(p)]] <- core.taxa
}

# Create a binary matrix indicating the presence (1) or absence (0) of genera in the core microbiome
core_matrix <- matrix(0, nrow = length(genus_list), ncol = length(core_genus_list),
                      dimnames = list(genus_list, names(core_genus_list)))

for (p in names(core_genus_list)) {
  core_matrix[core_genus_list[[p]], p] <- 1
}

# Remove rows and columns that sum to zero
core_matrix <- core_matrix[rowSums(core_matrix) > 0, colSums(core_matrix) > 0]

# Replace OTU names with genus names in the tax_table
tax_table_genus <- as.data.frame(tax_table(core_ps_genus))
genus_names <- tax_table_genus[rownames(core_matrix), "Genus"]

# Create a new matrix with genus names as rownames
core_matrix_genus <- core_matrix
genus_names <- gsub("g__", "", genus_names)

rownames(core_matrix_genus) <- genus_names


# Sort the matrix by row sums
row_sums <- rowSums(core_matrix_genus)
core_matrix_genus <- core_matrix_genus[order(row_sums, decreasing = TRUE), ]



# Convert the data frame to row-wise format while keeping row names
core_df_genus <- core_matrix_genus %>% as.data.frame() %>% rownames_to_column(var="Genera") %>%
  rowwise() %>%
  mutate(row_total = sum(c_across(starts_with("0.")))) %>%
  arrange(dplyr::desc(row_total)) %>%
  ungroup() %>%
  mutate(Genera = factor(Genera, levels = unique(Genera))) %>%
  dplyr::select(-row_total) %>% 
  pivot_longer(cols = -Genera,
               names_to = "prevalence",
               values_to = "value") 

# Pivoting the data longer and plotting
plot <- ggplot(core_df_genus, aes(x = prevalence, y = Genera, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "grey", name = "Prevalence") +
  labs(title = "Core Genera at Different Prevalence Levels",
       x = "Prevalence Level",
       y = "Genera") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8)
  )
plot
#ggsave(plot, filename = "~/Desktop/core_genera.png", width = 15, height = 25, units = "cm")
```


# Differential Abundance
## 6.1 ANCOMBC2

### 6.1.1 Phylum
```{r, fig.height=7, fig.width=10}
runANCOM(ps_obj_ped_fs, 
         tax_level = "Phylum", 
         group = "Type", 
         name_of_saved_results = "ancombc_results_phy", 
         plot_type = "dot",
         Log2FC_cutoff = 0.5,
         plot_heights = c(1,2),
         plot_colors = plot_colors )
```

### 6.1.2 Genus
```{r, fig.height=30, fig.width=10}
runANCOM(ps_obj_ped_fs, 
         tax_level = "Genus", 
         group = "Type", 
         name_of_saved_results = "ancombc_results_gen", 
         plot_type = "dot",
         plot_heights = c(1, 8),
         plot_colors = plot_colors )
```

### 6.1.3 Species
```{r, fig.height=45, fig.width=10}
runANCOM(ps_obj_ped_fs, 
         tax_level = "Species", 
         group = "Type", 
         name_of_saved_results = "ancombc_results_sp", 
         plot_type = "dot",
         #Log2FC_cutoff = 0.5,
         plot_heights = c(1, 8),
         plot_colors = plot_colors )
```


## 6.2 Maaslin2

### 6.2.1 Iterate Parameters
```{r, eval=FALSE, fig.height=8, fig.width=10}
resolutions <- c("Species", "Genus", "Phylum")
iterate_maaslin2(ps_obj = ps_obj_ped_fs, 
                 iterative_methods = iterative_methods,
                 resolutions = resolutions,
                 group = "condition",
                 qval_threshold = .25, 
                 plot_colors = maaslin2_colors, 
                 percentage = TRUE)
```

### 6.2.2 Run Maaslin2

#### Genus
```{r, fig.width=10, fig.height=65}
run_Maaslin2(ps_obj = ps_obj_ped_fs, 
                                 taxa_level = "Genus", 
                                 group = "condition",
                                 analysis_method = "LM", 
                                 normalization = "CSS",
                                 transform = "LOG", 
                                 plot_colors = plot_colors,
                                 plot_type = "dot", 
                                 qval_threshold= 0.1, 
                                 plot_heights = c(1, 9))
```

#### Species
```{r, fig.width=10, fig.height=65}
run_Maaslin2(ps_obj = ps_obj_ped_fs, 
                                 taxa_level = "Species", 
                                 group = "condition",
                                 analysis_method = "LM", 
                                 normalization = "CSS",
                                 transform = "LOG", 
                                 plot_colors = plot_colors,
                                 plot_type = "dot", 
                                 qval_threshold= 0.1, 
                                 plot_heights = c(1, 9))
```

## 6.3 AlDEX2

### 6.3.1 Iterate Parameters
https://www.nature.com/articles/s41467-022-28034-z

"...performed CLR transformation of each realization, and then performed Wilcoxon tests on the transformed realizations. The function then returned the expected Benjamini-Hochberg (BH) FDR-corrected p-value for each feature based on the results the different across Monte Carlo samples."

ALDEx2 uses by default the centred log-ratio (clr) transformation which is scale invariant

```{r, eval=FALSE,  fig.width=20, fig.height=12}
resolutions <- c("Species", "Genus", "Phylum")

 
iterate_aldex2(ps_obj = ps_obj_ped_fs, 
                 iterative_methods = iterative_methods,
                 resolutions = resolutions,
                 group = "condition",
                 percentage = FALSE)

aldex2_iterative_summary_table %>%
  filter(resolution == "Genus") %>%
  filter(paired == FALSE) %>%
  filter(paired == FALSE) %>%
  filter(denom == "all") %>%
  filter(transform == "log10") %>%
  dplyr::arrange(desc(significant_features)) %>%
  filter(significant_features > 0)

```

### 6.3.2 Run Aldex2

#### Genus
```{r}
run_aldex2(ps_obj_ped_fs, 
                       "condition", 
                       "Genus",
                       method = "wilcox.test",
                       transform = "log10",
                       normalization = "CSS")
```

#### Species
```{r, fig.width=12}
run_aldex2(ps_obj_ped_fs, 
                       "condition", 
                       "Species",
                       method = "wilcox.test",
                       transform = "log10",
                       normalization = "CSS")
```


## 6.4 Combine results

### Genus
```{r, fig.height=6, fig.width=5}
combine_DA(maaslin2_results_Genus, ancombc2_results_df_Genus, aldex2_res_Genus, "Type", "Genus", c(1,3))
```

### Species
```{r, fig.height=10, fig.width=5}
combine_DA(maaslin2_results_Species, ancombc2_results_df_Species, aldex2_res_Species, "Type", "Species", c(1,4))
```




# 8.0 Topic Modeling

## Genus .001

### Prep and binarize input
```{r, fig.width=4, fig.height=7}
#Prepare data from phyloseq object and binarize at a certain threshold of relative abundance
prep_data_binarize <- function(phy_obj, taxa_level, binarization_threshold){
  # Extract metadata
  meta_data <- phy_obj@sam_data %>% 
    as.matrix() %>% as.data.frame() %>% 
    dplyr::select(c("Type", "Sample"))
  
  #Make counts table - not currently used, but saving this
  #counts_data <- tax_glom(phy_obj, taxrank = taxa_level, NArm = TRUE) %>%
  #  psmelt() %>% as.matrix() %>% as.data.frame() %>%
  #  dplyr::select(c("Sample", "Abundance", taxa_level)) %>%
  #  pivot_wider(names_from = taxa_level, values_from = "Abundance") %>%
  #  column_to_rownames(var = "Sample")
  
  #Create Relative Abundance Table
  counts_data <- tax_glom(phy_obj, taxa_level) %>%
    transform_sample_counts(., function(x) (x + 0.000001 - min(x))) %>%
    transform_sample_counts(., function(x) (x) / sum(x)) %>% psmelt(.)  %>%
    dplyr::select(c("Sample", "Abundance", taxa_level)) %>%
    pivot_wider(names_from = taxa_level, values_from = "Abundance") %>%
    column_to_rownames(var="Sample") %>%
    mutate(across(where(is.numeric), ~ ifelse(. > binarization_threshold, 1, 0))) #Apply binarization threshold

  # Extract row names
  row_names <- rownames(counts_data)
  
  # Convert columns to integer
  counts_data <- as.data.frame(lapply(counts_data, as.integer))
  
  # Reassign row names
  rownames(counts_data) <- row_names
  

  #Reformat data into long format
  counts_data_long <- counts_data %>%
    pivot_longer(
      cols = everything(), 
      names_to = "taxa", 
      values_to = "Count"
    )
  
  #Create a histogram
  p <- ggplot(counts_data_long, aes(x = Count)) +
    geom_histogram(binwidth = 0.5, fill = "blue", color = "black", alpha = 0.7) +
    labs(title = "Histogram of Values", x = "Value", y = "Frequency") +
    stat_bin(
        binwidth = 0.5,
        aes(label = after_stat(count)),
        geom = "text",
        vjust = -0.5, # Adjust the vertical position of the text
        color = "black",
        size = 3
    ) +
    theme_bw()
  
  print(p)
  
  
  # Return metadata and counts data as a named list
  return(list(meta_data = meta_data, counts_data = counts_data))
  
}

results_threshold <- prep_data_binarize(ps_obj_ped_fs, "Genus", .001)
meta_data <- results_threshold$meta_data

counts_data <- results_threshold$counts_data
```

### Use FindTopicsNumber() 
Calculates different metrics to estimate the most preferable number of topics for LDA model.
CaoJuan2009: https://www.sciencedirect.com/science/article/pii/S092523120800372X
Arun2010: https://link.springer.com/chapter/10.1007/978-3-642-13657-3_43

Questions: How does this relate to perplexity? What about alpha values?
```{r, fig.width=10, fig.height=5}

RunFindTopicsNumber <- function(counts_data, topics, method){ 
  result <- FindTopicsNumber(
      counts_data,
      topics = topics,
      metrics = c("Griffiths2004", "CaoJuan2009", "Arun2010", "Deveaud2014"),
      method = method,
      control = list(seed = 1234),
      mc.cores = 2L,
      verbose = TRUE
    ) 
  FindTopicsNumber_plot(result)
}

topics <- seq(from = 2, to = 50, by = 2)
RunFindTopicsNumber(counts_data, topics, "Gibbs")

```

### Iteratve over alpha and k-values
Lower perplexity score = better
Alpha values: alpha represents document-topic density - with a higher alpha, documents are made up of more topics, and with lower alpha, documents contain fewer topics. 
Beta represents topic-word density - with a high beta, topics are made up of most of the words in the corpus, and with a low beta they consist of few words.
```{r, fig.height=12, fig.width=12}
iterate_alpha_kvalues <- function(counts_data, topic_no_iterations, alpha_iterations, method){
  
    # Create an empty dataframe to store results
  perplexity_alpha_results <- data.frame(
    k_value = integer(),
    alpha = numeric(),
    perplexity = numeric(),
    stringsAsFactors = FALSE
  )
      
  # Generate all combinations using expand.grid
      combinations <- expand.grid(topic_no = topic_no_iterations,
                                  alpha = alpha_iterations)
      
      # Convert each row into a list and store in a list
      iterative_methods <- apply(combinations, 1, function(x) {
        list(topic_no = x["topic_no"],
             alpha = x["alpha"])
        })
        
      for (params in iterative_methods) {
        
        # Extract parameters for the current iteration
        k_value <- as.numeric(params$topic_no)
        alpha <- as.numeric(params$alpha)

        lda_model <- LDA(counts_data, k = k_value, method = method, control = list(seed = 1234, alpha=alpha))
        perplex = perplexity(lda_model, newdata=as.matrix(counts_data))
        
      perplexity_alpha_results <- rbind(
      perplexity_alpha_results,
      data.frame(k_value = k_value, alpha = alpha, perplexity = perplex)
    )
  }

        # Assuming `results` is the dataframe returned by your function
    P <- ggplot(perplexity_alpha_results, aes(x = alpha, y = perplexity)) +
      geom_line() +  # Line for perplexity
      geom_point() + # Points for clarity
      facet_wrap(~ k_value, ncol=3) +  # Facet by k_value
      labs(
        title = "Perplexity Across Alpha Values for Different k_values",
        x = "Alpha Value",
        y = "Perplexity"
      ) +
        theme_bw() +
        theme(
          strip.text = element_text(size = 12, face = "bold"),  # Facet labels
          axis.text = element_text(size = 10),
          axis.title = element_text(size = 12),
          plot.title = element_text(size = 14, face = "bold")
        )
    
    P
}

topic_no_iterations <- seq(2, 20, by=2)
alpha_iterations <- seq(1, 100, by=2)
iterate_alpha_kvalues(counts_data, topic_no_iterations, alpha_iterations, "Gibbs")


```



### Create the model

https://bookdown.org/josephine_lukito/j381m_tutorial/ldatm.html
```{r}

#CREATE THE MODEL
create_topic_model <- function(counts_data, k_value, alpha, method){
  
    lda_model <- LDA(counts_data, k = k_value, method = method, control = list(seed = 1234, alpha=alpha))
    print(paste0("Perplexity is: ", perplexity(lda_model, newdata=as.matrix(counts_data))))
    
    b_df <- data.frame(tidytext::tidy(lda_model, matrix = "beta"))
    g_df <- data.frame(tidytext::tidy(lda_model, matrix = "gamma")) %>%
        arrange(document, topic)
    print(paste0("Alpha value is: ", lda_model@alpha))
  
    return(list(lda_model = lda_model, b_df = b_df, g_df=g_df))
}

result <- create_topic_model(counts_data, 6, 1, "Gibbs")


```

### Plot beta
```{r, fig.height=7, fig.width=10}
#Plot beta, or the numbers that are assigned to each word in a topic. If the beta score is higher, the word matters more in that topic.

plot_beta <- function(lda_result){
    top_terms <- result$b_df %>% 
      group_by(topic) %>% #groups by topic
      top_n(10, beta) %>% #takes the words with the top 10 beta scores
      ungroup() #ungroups the topic
    
    top_terms %>% 
      ggplot(aes(x = tidytext::reorder_within(term, beta, topic, sep = "_"), 
                 y =  beta, 
                 fill = factor(topic))) +
      geom_bar(stat = 'identity', show.legend = FALSE) +
      facet_wrap(~ topic, scales = "free") + 
      coord_flip()

}

plot_beta(results)
```
### Heatmap of gamma scores 
```{r}
#If you view the topics_wide data frame, you can see that each document has a gamma score for each topic. Some gamma scores are larger than others. This suggests that a document’s content is predominantly in one topic as opposed to another.

heatmap_gamma <- function(lda_results){
    topics_wide <- result$g_df %>%
      pivot_wider(names_from = topic,
                  values_from = gamma)

    topics_wide_type <- meta_data%>% select(c("Type")) %>% 
      rownames_to_column(var="document") %>% 
      merge.data.frame(topics_wide, by="document")
    topics_wide_type <- topics_wide_type[order(topics_wide_type$Type), ]
    
    annotations <- topics_wide_type%>% select(c( "document", "Type")) %>%
      mutate(Type = factor(Type, levels = c("Plaque", "Abscess"))) %>% # Reorder levels
      remove_rownames() %>% column_to_rownames(var="document")
        
    data <- topics_wide_type %>% select(-c("Type")) %>% remove_rownames() %>% column_to_rownames(var="document") %>% t()
        
    # Create the heatmap
    pheatmap(
          data,
          annotation_col = annotations,  # Add column annotations
          scale = "row",               # Scale rows (optional)
          cluster_rows = TRUE,         # Cluster rows
          cluster_cols = FALSE,         # Cluster columns
          color = colorRampPalette(c("blue", "white", "red"))(50),  # Custom color palette
          show_rownames = TRUE,        # Show row names
          show_colnames = FALSE         # Show column names
        )
}

heatmap_gamma(results)
```




### UMAP of gamma scores
```{r}
library(umap)

plot_gamma_umap <- function(lda_results){
    topics_wide <- result$g_df %>%
          pivot_wider(names_from = topic,
                      values_from = gamma) 
    
    res_with_annotations <- meta_data %>% select(c("Type")) %>% 
          rownames_to_column(var="document") %>% 
          merge.data.frame(topics_wide, by="document") 
    
    #Set UMAP seed
    custom.config <- umap.defaults
    custom.config$random_state <- 1234
    
    # Run UMAP
    umap_result <- res_with_annotations %>% select(-c("Type")) %>% column_to_rownames(var="document") %>% umap(., config=custom.config)
    
    # Convert the annotations to numeric for coloring
    annotations <- res_with_annotations %>% column_to_rownames(var="document") %>%  pull(Type)  # Extract "Type" as a vector
    type_colors <- c("Abscess" = "red", "Plaque" = "blue")  # Assign colors to each type
    colors <- type_colors[annotations]  # Match colors to each sample
    
    # Plot UMAP
    plot(umap_result$layout, 
         col = colors,
         pch = 19,           # Point type
         xlab = "UMAP 1", 
         ylab = "UMAP 2", 
         main = "UMAP Visualization")
}

plot_gamma_umap(results)
```



### Run model and plot
```{r, fig.width=8, fig.height=6}
result <- create_topic_model(counts_data, 6, 1, "Gibbs")

plot_beta(results)
heatmap_gamma(results)
plot_gamma_umap(results)
```

