---
title: "02a Pediatric"
author: "Madeline Krieger"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    theme: "cosmo"
    code_folding: "show"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE,
                      error = FALSE,
                      fig.path = "./figures/02a/",
                      fig.keep = "all",
                      dev = c("png"),
                      cache = FALSE,
                      # Do not use lazy loading which fails for large objects
                      cache.lazy = FALSE)
```


# 1.0 Load and Define

## 1.1 Load packages and presets
### 1.1.1 Packages
```{r 1.1_load-packages}
library("pacman")
pacman::p_load("rmarkdown", "tidyverse", "phyloseq", "knitr","kableExtra", "here", "plyr", "ggpubr", "microViz", "readxl", "microbiome", "pheatmap", "vegan", "reshape2", "magrittr", "microshades", "pheatmap","vegan", "data.table", "Polychrome", "fantaxtic","cetcolor", "ldatuning", "cowplot", "MicrobiomeStat", "randomForest", "caret", "mlbench", "MLmetrics", "mia", "here", "patchwork", "digest", "ANCOMBC", "Maaslin2", "microbiomeMarker")

here::i_am("02a_Pediatric.Rmd")

``` 


### 1.1.2 My packages
```{r}
load("MK_Microbiome_Functions.RData")
```



### 1.1.2 Colors and presents
```{r}
colors_all= c("Abscess" = "#FF495C", "Tumor"="#F0C808","Plaque"="#083D77", "AN"="#43AA8B" )
plot_colors= c("Abscess" = "#FF495C", "Plaque"="#083D77","abscess" = "#FF495C", "plaque"="#083D77", 
               "abnormal" = "#FF495C", "control"="#083D77",  "Abnormal" = "#FF495C", "Control"="#083D77"  )


core_colors = c("#FAAA00", "#3399FF","#F76F8E","#083D77","#B8D4E3", "#FF495C","#477071", '#03CEA4',  "#5F00BA", "#BDAC9E", "white", "#FFD900")

maaslin2_colors= c("CLR_LOG" = "#2c45b5", "CLR_NONE"="#86B8FD", 
                   "CSS_LOG" = "#FF495C", "CSS_NONE"="#F2929A",
                   "TMM_LOG" = "#F4B701", "TMM_NONE"="#FEE59A",
                   "TSS_LOG" = "#8FD694", "TSS_NONE"="#D2EED4",
                   "TSS_LOGIT" = "#354F52", "TSS_AST"="#CEDDDF")

pal20 <- createPalette(22, c("#F76F8E","#03CEA4", "#083D77"), range = c(30, 80))
pal20 <- unname(as.list(pal20))

resolution <- c("Phylum", "Class", "Order", "Family", "Genus", "Species")

set.seed(1234)
date = Sys.Date()
```


## 1.2 Load data and process

### 1.2.1 Load Data from 00
```{r}
# Load the saved Rds file
loaded_objects <- readRDS("00data.rds")

# Assign back the objects to the environment, if needed
list2env(loaded_objects, .GlobalEnv)

ps_obj_ped_f <- subset_samples(merged_phylo_obj_f, Study == "Pediatric")
ps_obj_ped_fs <- subset_samples(merged_phylo_obj_fs, Study == "Pediatric")
```


### 1.2.2 Process Phyloseq
```{r}
process_phyloseq(ps_obj_ped_fs)
```


# 2.0 Ordinations

## 2.1 PCA 
```{r, fig.width=10, fig.height=15}
plots <- lapply(resolution, function(rank) {
  plot_pca(phyloseq_obj = ps_obj_ped_f, 
           rank_transformation = rank, 
           variable = "Type", 
           colors_list = colors_all)})

combined_plot <- wrap_plots(plots, ncol = 2) & theme(legend.position = "bottom") 
combined_plot
```


## 2.2 PCoA NMDS Bray 
```{r, fig.width=10, fig.height=15}
plots <- lapply(resolution, function(rank) {
  plot_PCoA(phyloseq_obj = ps_obj_ped_f,
    rank_transformation = rank,
    trans_type = "identity",        
    dist_cal_type = "bray",   
    ord_calc_method = "NMDS",
    variable = "Type", 
    colors_list = colors_all)})

combined_plot <- wrap_plots(plots, ncol = 2) & theme(legend.position = "bottom") 
print(combined_plot)
```


## 2.3 PCoA NMDS Jaccard 
```{r, fig.width=10, fig.height=15}
plots <- lapply(resolution, function(rank) {
  plot_PCoA(phyloseq_obj = ps_obj_ped_f,
    rank_transformation = rank,
    trans_type = "identity",        
    dist_cal_type = "jaccard",   
    ord_calc_method = "NMDS",
        variable = "Type", 
    colors_list = colors_all)})

combined_plot <- wrap_plots(plots, ncol = 2) & theme(legend.position = "bottom") 
print(combined_plot)
```



# 3.0 Alpha diversity
```{r}
a_my_comparisons <- list( c("Abscess", "Plaque"))
symnum.args = list(cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1), symbols = c("****", "***", "**", "*", "ns"))

p_Shannon <- plot_richness(ps_obj_ped_f, x="Type", measures="Shannon", color = "Type")+
  geom_boxplot(alpha=0.6)+ 
    theme_bw(base_size=14) +
  theme(legend.position="none",         
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12), 
        axis.title.x = element_blank(),
        axis.title.y = element_text(size=12), 
        strip.background = element_rect(fill = "white", color = "white"), 
        strip.text = element_text(size = 14, face = "bold"))  +
  stat_compare_means(method = "wilcox.test", comparisons = a_my_comparisons, label = "p.format")+
  scale_color_manual(values = colors_all) +
  ylim(0, 6.2)


p_Chao1 <- plot_richness(ps_obj_ped_f, x="Type", measures="Chao1", color = "Type")+
  geom_boxplot(alpha=0.6)+ 
    theme_bw(base_size=14) +
  theme(legend.position="none",         
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12), 
        axis.title.x = element_blank(),
        axis.title.y = element_text(size=12), 
        strip.background = element_rect(fill = "white", color = "white"), 
        strip.text = element_text(size = 14, face = "bold"))  +
  stat_compare_means(method = "wilcox.test", comparisons = a_my_comparisons, label = "p.format")+
  scale_color_manual(values = colors_all) +
  ylim(0, 470)


p_Observed <- plot_richness(ps_obj_ped_f, x="Type", measures="Observed", color = "Type")+
  geom_boxplot(alpha=0.6)+ 
    theme_bw(base_size=14) +
  theme(legend.position="none",         
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12), 
        axis.title.x = element_blank(),
        axis.title.y = element_text(size=12), 
        strip.background = element_rect(fill = "white", color = "white"), 
        strip.text = element_text(size = 14, face = "bold"))  +
  stat_compare_means(method = "wilcox.test", comparisons = a_my_comparisons, label = "p.format")+
  scale_color_manual(values = colors_all) +
  ylim(0, 470)

plot <- ggarrange(p_Shannon, p_Observed, p_Chao1, ncol=3, nrow=1, common.legend = TRUE, legend="bottom")
annotate_figure(plot, top = text_grob(paste0("Alpha Diversity"), color = "black", face = "bold", size = 14))
```

# 4.0 Taxa Plots

## 4.1 Average Bar Plots

### 4.1.1 Phylum 
```{r taxa_phylum, fig.height=4, fig.width=4}

top_df <- ps_obj_ped_fs_t_n %>%
  tax_glom(., "Phylum") %>%
  get_top_taxa(., 10, relative = TRUE, discard_other = TRUE) %>% psmelt(.)
top_df$Phylum <- gsub("p__","",as.character(top_df$Phylum))

top_df %>%
  dplyr::group_by(Type,Phylum)%>%
  dplyr::summarise(Average_Abundance = mean(Abundance))%>%
  ggbarplot(x= "Type", y="Average_Abundance",
            fill = "Phylum",
            panel.labs.font = list(size=12),
            panel.labs.background = list(color = NULL, fill = "white"),
            font.tickslab = 14,
            font.legend = c(10),
            palette = core_colors, 
            strip.position = "top",
            color = "black",
            x.text.angle = 45, 
            y.text.angle = 0, 
            font.x = 14,
            font.y = 12,
            font.title = 16,
            rotate = FALSE,
            title = "Phlyum Level", 
            xlab = "",
            ylab = "Average Relative Abundance",
            ggtheme = theme_bw()) + 
  font("title", face="bold") +
  theme(plot.title=element_text(hjust=0.5), plot.subtitle=element_text(hjust=0.5))
```

### 4.1.2 Genus 
```{r taxa_genus, fig.height=4, fig.width=3}

top_df <- ps_obj_ped_fs_t_n %>%
  tax_glom(., "Genus") %>%
  get_top_taxa(., 10, relative = TRUE, discard_other = TRUE) %>% psmelt(.)
top_df$Genus <- gsub("g__","",as.character(top_df$Genus))

top_df %>%
  dplyr::group_by(Type,Genus)%>%
  dplyr::summarise(Average_Abundance = mean(Abundance))%>%
  ggbarplot(x= "Type", y="Average_Abundance",
            fill = "Genus",
            panel.labs.font = list(size=12),
            panel.labs.background = list(color = NULL, fill = "white"),
            font.tickslab = 14,
            font.legend = c(10),
            palette = core_colors, 
            strip.position = "top",
            color = "black",
            x.text.angle = 45, 
            y.text.angle = 0, 
            font.x = 14,
            font.y = 12,
            font.title = 16,
            rotate = FALSE,
            title = "Top 10 Genera", 
            xlab = "",
            ylab = "Average Relative Abundance",
            ggtheme = theme_bw()) + 
  font("title", face="bold") +
  theme(plot.title=element_text(hjust=0.5), plot.subtitle=element_text(hjust=0.5), legend.text = element_text(face = "italic"))
```

### 4.1.3 Microshades Genus MB
```{r, fig.width=8, fig.height=4}
# Use microshades function prep_mdf to agglomerate, normalize, and melt the phyloseq object
mdf_prep <- prep_mdf(ps_obj_ped_fs)

mdf_prep <- mdf_prep %>%
  mutate(Kingdom = gsub("k__", "", Kingdom)) %>%
  mutate(Phylum = gsub("p__", "", Phylum)) %>%
  mutate(Class = gsub("c__", "", Class)) %>%
  mutate(Order = gsub("o__", "", Order)) %>%
  mutate(Family = gsub("f__", "", Family)) %>%
  mutate(Genus = gsub("g__", "", Genus)) %>%
  mutate(Species = gsub("s__", "", Species)) 

print(unique(mdf_prep$Phylum))

# Create a color object for the specified data
color_obj <- create_color_dfs(mdf_prep, 
            selected_groups = c("Proteobacteria", "Actinobacteria","Bacteroidetes", "Fusobacteria",
                                "Spirochaetes"),
    cvd = FALSE, top_orientation=TRUE)

# Extract
mdf <- color_obj$mdf
cdf <- color_obj$cdf

plot_1 <- plot_microshades(mdf, cdf)

plot_1 + scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme_bw() +
  theme(legend.key.size = unit(0.2, "cm"), text=element_text(size=10), ) +
  theme(axis.text.x = element_text(size= 6)) +
  facet_wrap(~Type, scales= "free_x") 
```



### 4.1.4 Species 
```{r species}
top_df <- ps_obj_ped_fs_t_n %>%
  tax_glom(., "Species") %>%
  get_top_taxa(., 10, relative = TRUE, discard_other = TRUE) %>% psmelt(.)
top_df$Species <- gsub("s__","",as.character(top_df$Species))

top_df %>%
  dplyr::group_by(Type,Species)%>%
  dplyr::summarise(Average_Abundance = mean(Abundance))%>%
  ggbarplot(x= "Type", y="Average_Abundance",
            fill = "Species",
            panel.labs.font = list(size=12),
            panel.labs.background = list(color = NULL, fill = "white"),
            font.tickslab = 10,
            font.legend = c(10),
            palette = pal20, 
            strip.position = "top",
            color = "black",
            x.text.angle = 90, 
            y.text.angle = 0, 
            font.x = 12,
            font.y = 12,
            font.title = 16,
            rotate = TRUE,
            title = "", 
            subtitle = "Species level",
            xlab = "Condition",
            ylab = "Average Abundance",
            ggtheme = theme_bw()) + 
  font("title", face="bold") +
  theme(plot.title=element_text(hjust=0.5), plot.subtitle=element_text(hjust=0.5))
```


## 4.2 Per taxa plots

### 4.2.1 Genus TSS Dot
```{r, fig.width=15, fig.height=50}
plot_all_taxa(ps_obj_ped_fs, "TSS", "condition", "Genus", "dot", plot_colors)
```


# 5.0 Core Microbiome

Original code below, I'm not 100% certain it's correct
```{r, fig.height=9, fig.width=7, eval=FALSE}

# Get the total list of species
genus_list <- taxa_names(core_ps_genus) 

# Initialize an empty list to store core genera at each prevalence level
core_genus_list <- list()

# Loop through prevalence values from 0.1 to 1
for (p in seq(0.1, 1, by = 0.05)) {
  core.taxa <- core_members(core_ps_genus, detection = 0, prevalence = p)
  core_genus_list[[as.character(p)]] <- core.taxa
}

# Create a binary matrix indicating the presence (1) or absence (0) of genera in the core microbiome
core_matrix <- matrix(0, nrow = length(genus_list), ncol = length(core_genus_list),
                      dimnames = list(genus_list, names(core_genus_list)))

for (p in names(core_genus_list)) {
  core_matrix[core_genus_list[[p]], p] <- 1
}

# Remove rows and columns that sum to zero
core_matrix <- core_matrix[rowSums(core_matrix) > 0, colSums(core_matrix) > 0]

# Replace OTU names with genus names in the tax_table
tax_table_genus <- as.data.frame(tax_table(core_ps_genus))
genus_names <- tax_table_genus[rownames(core_matrix), "Genus"]

# Create a new matrix with genus names as rownames
core_matrix_genus <- core_matrix
genus_names <- gsub("g__", "", genus_names)

rownames(core_matrix_genus) <- genus_names


# Sort the matrix by row sums
row_sums <- rowSums(core_matrix_genus)
core_matrix_genus <- core_matrix_genus[order(row_sums, decreasing = TRUE), ]

# Convert the data frame to row-wise format while keeping row names
core_df_genus <- core_matrix_genus %>% as.data.frame() %>% rownames_to_column(var="Genera") %>%
  rowwise() %>%
  mutate(row_total = sum(c_across(starts_with("0.")))) %>%
  arrange(dplyr::desc(row_total)) %>%
  ungroup() %>%
  mutate(Genera = factor(Genera, levels = unique(Genera))) %>%
  dplyr::select(-row_total) %>% 
  pivot_longer(cols = -Genera,
               names_to = "prevalence",
               values_to = "value") 

# Pivoting the data longer and plotting
plot <- ggplot(core_df_genus, aes(x = prevalence, y = Genera, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "grey", name = "Prevalence") +
  labs(title = "Core Genera at Different Prevalence Levels",
       x = "Prevalence Level",
       y = "Genera") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8)
  )
plot
```


# Differential Abundance
## 6.1 ANCOMBC2

### 6.1.1 Phylum
```{r, fig.height=7, fig.width=10}
runANCOM(ps_obj_ped_fs, 
         tax_level = "Phylum", 
         group = "Type", 
         name_of_saved_results = "ancombc_results_phy", 
         plot_type = "dot",
         Log2FC_cutoff = 0.5,
         plot_heights = c(1,2),
         plot_colors = plot_colors )
```

### 6.1.2 Genus
```{r, fig.height=30, fig.width=10}
runANCOM(ps_obj_ped_fs, 
         tax_level = "Genus", 
         group = "Type", 
         name_of_saved_results = "ancombc_results_gen", 
         plot_type = "dot",
         plot_heights = c(1, 8),
         plot_colors = plot_colors )
```

### 6.1.3 Species
```{r, fig.height=45, fig.width=10}
runANCOM(ps_obj_ped_fs, 
         tax_level = "Species", 
         group = "Type", 
         name_of_saved_results = "ancombc_results_sp", 
         plot_type = "dot",
         #Log2FC_cutoff = 0.5,
         plot_heights = c(1, 8),
         plot_colors = plot_colors )
```


## 6.2 Maaslin2

### 6.2.1 Iterate Parameters
```{r, eval=FALSE, fig.height=8, fig.width=10}
resolutions <- c("Species", "Genus", "Phylum")
iterate_maaslin2(ps_obj = ps_obj_ped_fs, 
                 iterative_methods = iterative_methods,
                 resolutions = resolutions,
                 group = "condition",
                 qval_threshold = .25, 
                 plot_colors = maaslin2_colors, 
                 percentage = TRUE)
```

### 6.2.2 Run Maaslin2

#### Genus
```{r, fig.width=10, fig.height=65}
run_Maaslin2(ps_obj = ps_obj_ped_fs, 
                                 taxa_level = "Genus", 
                                 group = "condition",
                                 analysis_method = "LM", 
                                 normalization = "CSS",
                                 transform = "LOG", 
                                 plot_colors = plot_colors,
                                 plot_type = "dot", 
                                 qval_threshold= 0.25, 
                                 plot_heights = c(1, 9))
```

#### Species
```{r, fig.width=10, fig.height=65}
run_Maaslin2(ps_obj = ps_obj_ped_fs, 
                                 taxa_level = "Species", 
                                 group = "condition",
                                 analysis_method = "LM", 
                                 normalization = "CSS",
                                 transform = "LOG", 
                                 plot_colors = plot_colors,
                                 plot_type = "dot", 
                                 qval_threshold= 0.25, 
                                 plot_heights = c(1, 9))
```

## 6.3 AlDEX2

### 6.3.1 Iterate Parameters
https://www.nature.com/articles/s41467-022-28034-z

"...performed CLR transformation of each realization, and then performed Wilcoxon tests on the transformed realizations. The function then returned the expected Benjamini-Hochberg (BH) FDR-corrected p-value for each feature based on the results the different across Monte Carlo samples."

ALDEx2 uses by default the centred log-ratio (clr) transformation which is scale invariant

```{r, eval=FALSE,  fig.width=20, fig.height=12}
resolutions <- c("Species", "Genus", "Phylum")

 
iterate_aldex2(ps_obj = ps_obj_ped_fs, 
                 iterative_methods = iterative_methods,
                 resolutions = resolutions,
                 group = "condition",
                 percentage = FALSE)

aldex2_iterative_summary_table %>%
  filter(resolution == "Genus") %>%
  filter(paired == FALSE) %>%
  filter(paired == FALSE) %>%
  filter(denom == "all") %>%
  filter(transform == "log10") %>%
  dplyr::arrange(desc(significant_features)) %>%
  filter(significant_features > 0)

```

### 6.3.2 Run Aldex2

#### Genus
```{r}
run_aldex2(ps_obj_ped_fs, 
                       "condition", 
                       "Genus",
                       method = "wilcox.test",
                       transform = "log10",
                       normalization = "CSS")
```

#### Species
```{r, fig.width=12}
run_aldex2(ps_obj_ped_fs, 
                       "condition", 
                       "Species",
                       method = "wilcox.test",
                       transform = "log10",
                       normalization = "CSS")
```


## 6.4 Combine results

### Genus
```{r, fig.height=6, fig.width=5}
combine_DA(maaslin2_results_Genus, ancombc2_results_df_Genus, aldex2_res_Genus, "Type", "Genus", c(1,3))
```

### Species
```{r, fig.height=10, fig.width=5}
combine_DA(maaslin2_results_Species, ancombc2_results_df_Species, aldex2_res_Species, "Type", "Species", c(1,4))
```


