---
title: "03_TopicModeling"
author: "Madeline Krieger"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    theme: "cosmo"
    code_folding: "show"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE,
                      error = FALSE,
                      fig.path = "./figures/03/",
                      fig.keep = "all",
                      dev = c("png"),
                      cache = FALSE,
                      # Do not use lazy loading which fails for large objects
                      cache.lazy = FALSE)
```


# 1.0 Load and Define

## 1.1 Load packages and presets
### 1.1.1 Packages
```{r 1.1_load-packages}
library("pacman")
pacman::p_load("rmarkdown", "tidyverse", "phyloseq", "knitr","kableExtra", "here", "plyr", "ggpubr", "microViz", "readxl", "microbiome", "pheatmap", "vegan", "reshape2", "magrittr", "microshades", "pheatmap","vegan", "data.table", "Polychrome", "fantaxtic","cetcolor", "ldatuning", "cowplot", "MicrobiomeStat", "randomForest", "caret", "mlbench", "MLmetrics", "mia", "here", "patchwork", "digest", "ANCOMBC", "Maaslin2", "microbiomeMarker")

here::i_am("03_TopicModeling.Rmd")

``` 


### 1.1.2 My packages
```{r}
load("MK_Microbiome_Functions.RData")
```



### 1.1.2 Colors and presents
```{r}
colors_all= c("Abscess" = "#FF495C", "Tumor"="#F0C808","Plaque"="#083D77", "AN"="#43AA8B" )
plot_colors= c("Abscess" = "#FF495C", "Plaque"="#083D77","abscess" = "#FF495C", "plaque"="#083D77", 
               "abnormal" = "#FF495C", "control"="#083D77",  "Abnormal" = "#FF495C", "Control"="#083D77"  )

core_colors = c("#FAAA00", "#3399FF","#F76F8E","#083D77","#B8D4E3", "#FF495C","#477071", '#03CEA4',  "#5F00BA", "#BDAC9E", "white", "#FFD900")

maaslin2_colors= c("CLR_LOG" = "#2c45b5", "CLR_NONE"="#86B8FD", 
                   "CSS_LOG" = "#FF495C", "CSS_NONE"="#F2929A",
                   "TMM_LOG" = "#F4B701", "TMM_NONE"="#FEE59A",
                   "TSS_LOG" = "#8FD694", "TSS_NONE"="#D2EED4",
                   "TSS_LOGIT" = "#354F52", "TSS_AST"="#CEDDDF")

pal20 <- createPalette(22, c("#F76F8E","#03CEA4", "#083D77"), range = c(30, 80))
pal20 <- unname(as.list(pal20))

resolution <- c("Phylum", "Class", "Order", "Family", "Genus", "Species")

set.seed(1234)
date = Sys.Date()
```


## 1.2 Load data and process

### 1.2.1 Load Data from 00
```{r}
# Load the saved Rds file
loaded_objects <- readRDS("00data.rds")

# Assign back the objects to the environment, if needed
list2env(loaded_objects, .GlobalEnv)

ps_obj_ped_fs <- subset_samples(merged_phylo_obj_fs, Study == "Pediatric")
ps_obj_ped_adult_fs <- subset_samples(merged_phylo_obj_fs, Study != "OSCC")

```


# 2.0 Merged Ped and Adult

## 2.1 Genus

### Threshold iterations
```{r}
threshold_iterations <-  seq(.75, 1.02, by = .02)
result_table <- find_thresholds_iteratively(ps_obj_ped_adult_fs, "Genus", threshold_iterations)
result_table
```

### Find filtering threshold
```{r}
threshold <- find_threshold(ps_obj_ped_adult_fs, "Genus", .95)
threshold
```

### Prep and binarize input
```{r}
results_threshold <- prep_data_binarize(ps_obj_ped_adult_fs, "Genus", threshold, "condition")
meta_data <- results_threshold$meta_data
counts_data <- results_threshold$counts_data
```

### Use FindTopicsNumber() 
```{r}
topics <- seq(from = 2, to = 50, by = 2)
RunFindTopicsNumber(counts_data, topics, "Gibbs")
```

### LDA: Create the model
```{r}
result <- create_topic_model(counts_data, 2, 1, "Gibbs")
model <- result$lda_model
```

### Plot beta
```{r, fig.height=4, fig.width=10}
plot_beta(result, 15)
```

### Heatmap of gamma scores 
```{r, fig.height=2, fig.width=9}
heatmap_gamma(result, "condition")
```

### UMAP of gamma scores
```{r fig.height=4, fig.width=4}
plot_gamma_umap(results, "condition", plot_colors)
```

###Membership of topic by Type
```{r, fig.height=3.5, fig.width=4}
topic_membership(result, "condition", plot_colors)
```





## 2.2 Species

### Threshold iterations
```{r}
threshold_iterations <-  seq(.75, 1.02, by = .02)
result_table <- find_thresholds_iteratively(ps_obj_ped_adult_fs, "Genus", threshold_iterations)
result_table
```

### Find filtering threshold
```{r}
threshold <- find_threshold(ps_obj_ped_adult_fs, "Genus", .90)
threshold
```

### Prep and binarize input
```{r}
results_threshold <- prep_data_binarize(ps_obj_ped_adult_fs, "Genus", threshold, "condition")
meta_data <- results_threshold$meta_data
counts_data <- results_threshold$counts_data
```

### Use FindTopicsNumber() 
```{r}
topics <- seq(from = 2, to = 50, by = 2)
RunFindTopicsNumber(counts_data, topics, "Gibbs")
```

### LDA: Create the model
```{r}
result <- create_topic_model(counts_data, 2, 1, "Gibbs")
model <- result$lda_model
```

### Plot beta
```{r, fig.height=4, fig.width=10}
plot_beta(result, 15)
```

### Heatmap of gamma scores 
```{r, fig.height=2, fig.width=9}
heatmap_gamma(result, "condition")
```

### UMAP of gamma scores
```{r fig.height=4, fig.width=4}
plot_gamma_umap(results, "condition", plot_colors)
```

###Membership of topic by Type
```{r, fig.height=3.5, fig.width=4}
topic_membership(result, "condition", plot_colors)
```


# 3.0 Pediatric only

## Genus
### Threshold iterations
```{r}
threshold_iterations <-  seq(.75, 1.02, by = .02)
result_table <- find_thresholds_iteratively(ps_obj_ped_fs, "Genus", threshold_iterations)
result_table
```

### Find filtering threshold
```{r}
threshold <- find_threshold(ps_obj_ped_fs, "Genus", .90)
threshold
```

### Prep and binarize input
```{r}
results_threshold <- prep_data_binarize(ps_obj_ped_fs, "Genus", threshold, "condition")
meta_data <- results_threshold$meta_data
counts_data <- results_threshold$counts_data
```

### Use FindTopicsNumber() 
```{r}
topics <- seq(from = 2, to = 10, by = 1)
RunFindTopicsNumber(counts_data, topics, "Gibbs")
```

### LDA: Create the model
```{r}
result <- create_topic_model(counts_data, 2, 1, "Gibbs")
model <- result$lda_model
```

### Plot beta
```{r, fig.height=4, fig.width=10}
plot_beta(result, 15)
```

### Heatmap of gamma scores 
```{r, fig.height=2, fig.width=9}
heatmap_gamma(result, "condition")
```

### UMAP of gamma scores
```{r fig.height=6, fig.width=6}
library(umap)
plot_gamma_umap(results, "condition", plot_colors)
```

###Membership of topic by Type
```{r, fig.height=3.5, fig.width=4}
topic_membership(result, "condition", plot_colors)
```


### Heatmap of rel abundance in original data of top taxa
```{r, fig.height=5, fig.width=7}
relab_heatmap(result, ps_obj_ped_fs, "Genus", "condition", 1, 10)
```


## Species
### Threshold iterations
```{r}
threshold_iterations <-  seq(.75, 1.02, by = .02)
result_table <- find_thresholds_iteratively(ps_obj_ped_fs, "Species", threshold_iterations)
result_table
```

### Find filtering threshold
```{r}
threshold <- find_threshold(ps_obj_ped_fs, "Species", .90)
threshold
```

### Prep and binarize input
```{r}
results_threshold <- prep_data_binarize(ps_obj_ped_fs, "Species", threshold, "condition")
meta_data <- results_threshold$meta_data
counts_data <- results_threshold$counts_data
```

### Use FindTopicsNumber() 
```{r}
topics <- seq(from = 2, to = 50, by = 2)
RunFindTopicsNumber(counts_data, topics, "Gibbs")
```

### LDA: Create the model
```{r}
result <- create_topic_model(counts_data, 2, 1, "Gibbs")
model <- result$lda_model
```

### Plot beta
```{r, fig.height=4, fig.width=10}
plot_beta(result, 15)
```

### Heatmap of gamma scores 
```{r, fig.height=2, fig.width=9}
heatmap_gamma(result, "condition")
```

### UMAP of gamma scores
```{r fig.height=4, fig.width=4}
plot_gamma_umap(results, "condition", plot_colors)
```

###Membership of topic by Type
```{r, fig.height=3.5, fig.width=4}
topic_membership(result, "condition", plot_colors)
```


### Heatmap of rel abundance in original data of top taxa
```{r, fig.height=5, fig.width=7}
relab_heatmap(result, ps_obj_ped_fs, "Species", "condition", 1, 10)
```


